---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Start by loading the packages for data manipulation and plotting

```{r}
library(ggplot2)
library(scales)
library(plyr)
```

## Loading and preprocessing the data

1. Load the data (i.e. read.csv())

```{r}
unzip('activity.zip')
accel <- read.csv('activity.csv', header=TRUE, sep = ',')
str(accel)
for(i in 1:length(accel) ) {
  str(accel[i])
  }
```

2. Recode Factor date as a date variable.

```{r}
accel$date <- as.Date( accel$date, format = '%Y-%m-%d' )
str(accel$date)
```

## What is mean total number of steps taken per day?

1. Compute the data needed for the plot

```{r}
tot.steps <- ddply(accel, .(date),
                  summarize,
                  total.steps = sum(steps)
                  )
head(tot.steps)
```

2. Make a histogram of the total number of steps taken each day

```{r}
tot.st.plot <- ggplot(tot.steps,
                      aes(x = total.steps)
                      )
tot.st.plot <- tot.st.plot + geom_histogram()
print(tot.st.plot)
```

3. Report the summary statistics, to see the mean and median.

```{r}
summary(tot.steps$total.steps)
summary(tot.steps$total.steps)['Mean']
summary(tot.steps$total.steps)['Median']
```

## What is the average daily activity pattern?

1. Compute data

```{r}
av.steps <- ddply(accel, .(interval),
                  summarize,
                  average.steps = mean(steps, na.rm=TRUE
                                       )
                  )
head(av.steps)
```

2. Make a time series plot

```{r}
av.st.plot <- ggplot(av.steps,
                     aes(interval, average.steps)
                     )
av.st.plot <- av.st.plot + geom_line()
print(av.st.plot)
```

3. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
summary(av.steps$average.steps)
summary(av.steps$average.steps)['Max.']
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset

```{r}
table( complete.cases(accel) )
```

2. Create a new data frame and impute missing values using interval averages.

```{r}
imp.accel <- accel
imp.accel$steps <- ifelse( is.na(imp.accel$steps),
                     av.steps$average.steps[match(imp.accel$interval, av.steps$interval) ],
                     imp.accel$steps )
table( complete.cases(accel) )
table( complete.cases(imp.accel) )
```

```{r}
tot.imp.steps <- ddply(imp.accel, .(date),
                  summarize,
                  total.steps = sum(steps)
                  )
head(tot.imp.steps)
```

3. Make a histogram of the total number of steps taken each day

```{r}
tot.imp.st.plot <- ggplot(tot.imp.steps,
                      aes(x = total.steps)
                      )
tot.imp.st.plot <- tot.imp.st.plot + geom_histogram()
print(tot.imp.st.plot)
```

4. Calculate and report the mean and median total number of steps taken per day.

```{r}
summary(tot.imp.steps$total.steps)
summary(tot.imp.steps$total.steps)['Mean']
summary(tot.imp.steps$total.steps)['Median']
```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a day of the week variable in the accel data frame, then we create a list of weekdays and weekend days.

```{r}
accel$day <- weekdays(accel$date)

w.day <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
w.e.day <- c('Saturday', 'Sunday')

accel$weekpart <- ifelse(accel$day %in% w.day, 'weekday', 'weekend')
head(accel)
table(accel$weekpart)
```

2. Make a panel plot containing a time series plot

```{r}
av.w.e.steps <- ddply(accel, .(interval, weekpart),
                  summarize,
                  average.steps = mean(steps, na.rm=TRUE
                                       )
                  )
head(av.w.e.steps)
```

```{r}
av.w.e.st.plot <- ggplot(av.w.e.steps,
                     aes(interval, average.steps)
                     )
av.w.e.st.plot <- av.w.e.st.plot + geom_line()
av.w.e.st.plot <- av.w.e.st.plot + facet_grid(weekpart ~ .)
print(av.w.e.st.plot)
```



